---
title: Exploring sample overlap simulations (replication)
---

```{r}
suppressWarnings(suppressPackageStartupMessages({
  library(simulateGP)
  library(TwoSampleMR)
  library(dplyr)
  library(ggplot2)
  library(parallel)
  library(knitr)
  args <- commandArgs(T)
  imgdir <- args[1]
}))
opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE)
```

## Outline of simulation

1. Create x and y in a population
2. Find instruments for x in discovery and test for significance
3. Retest in replication
4. Extract those effects from outcome GWAS, which could be independent, discovery or replication sample



### Find GWAS parameters which will likely give high winner's curse


```{r}
simulate_gwasx <- function(nid, nsnp, bgx, bux)
{
  g <- make_geno(nid, nsnp, 0.5)
  u <- rnorm(nid)
  geneff <- rep(bgx, nsnp)
  x <- make_phen(c(geneff, bux), cbind(g, u))
  gwas(x, g)
}
sapply(1:1000, function(x)
{
  simulate_gwasx(3000, 20, 0.08, 0.4) %>%
    {sum(.$pval < 5e-8)}
}) %>% table
```


### Check that UMVCUE works at reducing winner's curse

```{r}
umvcue <- function(d_beta, r_beta, d_se, r_se, pthresh)
{
  t <- qnorm(1 - pthresh/2)
  MLE <- (r_se^2 * d_beta + d_se^2 * r_beta) / (r_se^2 + d_se^2)
  BB <- (sqrt(r_se^2 + d_se^2) / d_se^2) * (MLE - t * d_se)
  b <- MLE - (r_se^2 / sqrt(r_se^2 + d_se^2) * dnorm(BB) / pnorm(BB))
  return(b)
}
simulate_umvcue <- function(nid, nsnp, bgx, bux)
{
  geneff <- rep(bgx, nsnp)
  g <- make_geno(nid/2, nsnp, 0.5)
  u <- rnorm(nid/2)
  x <- make_phen(c(geneff, bux), cbind(g, u))
  d <- gwas(x, g)
  g <- make_geno(nid/2, nsnp, 0.5)
  u <- rnorm(nid/2)
  x <- make_phen(c(geneff, bux), cbind(g, u))
  r <- gwas(x, g)
  index <- d$pval < 5e-8
  dat <- tibble(
    truth = geneff,
    disc = d$bhat,
    rep = r$bhat,
    disc_sig = d$pval < 5e-8,
    what="all"
  )
  dat$umvcue <- NA
  if(any(index))
  {
    dat$umvcue[index] <- umvcue(d$bhat[index], r$bhat[index], d$se[index], r$se[index], 5e-8)
  }
  return(dat)
}
out <- lapply(1:1000, function(x)
{
  simulate_umvcue(3000, 20, 0.13, 0.4) %>%
    mutate(sim=x)
}) %>% bind_rows()
out2 <- group_by(out, sim) %>%
  summarise(
    nsig=sum(disc_sig), 
    truth=mean(truth),
    disc = mean(disc),
    rep = mean(rep),
    umvcue = mean(umvcue, na.rm=T)
  )
out3 <- filter(out, disc_sig) %>%
  group_by(sim) %>%
  summarise(
    nsig=sum(disc_sig), 
    truth=mean(truth),
    disc = mean(disc),
    rep = mean(rep),
    umvcue = mean(umvcue, na.rm=T)
  )
colMeans(out2)
colMeans(out3)
```

it does reduce winner's curse, gives simular result to replication


### Run MR simulations for different sampling strategies

```{r}
gwas_bias <- function(gwas)
{
  # sum((gwas$bhat - gwas$b)^2)
  mean(abs(gwas$bhat - gwas$b))
}
gwas_sim <- function(nid = 9000, nsnp = 20, bgx = 0.08, bxy = 0.2, buy = 0.4, bux = 0.4, out_index = 1:3000)
{
  # Simulate population
  g <- make_geno(nid, nsnp, 0.5)
  u <- rnorm(nid)
  geneff <- rep(bgx, nsnp)
  x <- make_phen(c(geneff, bux), cbind(g, u))
  y <- make_phen(c(buy, bxy), cbind(u, x))
  # Get observational estimate
  obs_beta <- summary(lm(y ~ x))$coefficients[2,1]
  # GWAS in samples
  disc_index <- 1:3000
  rep_index <- 3001:6000
  disc_gwas <- gwas(x[disc_index], g[disc_index,])
  rep_gwas <- gwas(x[rep_index], g[rep_index,])
  out_gwas <- gwas(y[out_index], g[out_index,])
  disc_gwas$b <- geneff
  rep_gwas$b <- geneff
  out_gwas$b <- geneff * bxy
  bias <- expand.grid(sample=c("disc", "rep", "out"), what=c("all", "sig"), bias=NA)
  bias$sse[1] <- gwas_bias(disc_gwas)
  bias$sse[2] <- gwas_bias(rep_gwas)
  bias$sse[3] <- gwas_bias(out_gwas)
  # what is significant
  index <- disc_gwas$pval < 5e-8
  bias$sse[4] <- gwas_bias(disc_gwas[index,])
  bias$sse[5] <- gwas_bias(rep_gwas[index,])
  bias$sse[6] <- gwas_bias(out_gwas[index,])
  bias$nsig <- sum(index)
  # Perform MR
  d <- list()
  d$disc_all <- simulateGP::make_dat(disc_gwas, out_gwas)
  d$rep_all <- simulateGP::make_dat(rep_gwas, out_gwas)
  d$disc_sig <- simulateGP::make_dat(disc_gwas[index,], out_gwas[index,])
  d$rep_sig <- simulateGP::make_dat(rep_gwas[index,], out_gwas[index,])
  d$umvcue_sig <- d$disc_sig
  d$umvcue_sig$beta.exposure <- umvcue(d$disc_sig$beta.exposure, d$rep_sig$beta.exposure, d$disc_sig$se.exposure, d$rep_sig$se.exposure, 5e-8)
  m <- lapply(names(d), function(x) {
    y <- d[[x]]
    if(nrow(y) > 0)
    {
      z <- mr(y, method_list=c("mr_ivw", "mr_wald_ratio")) %>% as_tibble()
      z$what <- x
      return(z)
    } else {
      return(NULL)
    }
  }) %>% bind_rows()
  return(list(bias=bias, mr=m))
}
res <- mclapply(1:1000, function(i) {
  message(i)
  bind_rows(
    gwas_sim(bxy=0.0, out_index = 1:3000, buy=-0.4)$mr %>% mutate(overlap="disc_overlap", bxy=0.0, buy=-0.4),
    gwas_sim(bxy=0.0, out_index = 3001:6000, buy=-0.4)$mr %>% mutate(overlap="rep_overlap", bxy=0.0, buy=-0.4),
    gwas_sim(bxy=0.0, out_index = 6001:9000, buy=-0.4)$mr %>% mutate(overlap="no_overlap", bxy=0.0, buy=-0.4),
    gwas_sim(bxy=0.0, out_index = 1501:4500, buy=-0.4)$mr %>% mutate(overlap="both_overlap", bxy=0.0, buy=-0.4),
    gwas_sim(bxy=0.0, out_index = 1:3000, buy=0)$mr %>% mutate(overlap="disc_overlap", bxy=0.0, buy=0),
    gwas_sim(bxy=0.0, out_index = 3001:6000, buy=0)$mr %>% mutate(overlap="rep_overlap", bxy=0.0, buy=0),
    gwas_sim(bxy=0.0, out_index = 6001:9000, buy=0)$mr %>% mutate(overlap="no_overlap", bxy=0.0, buy=0),
    gwas_sim(bxy=0.0, out_index = 1501:4500, buy=0)$mr %>% mutate(overlap="both_overlap", bxy=0.0, buy=0),
    gwas_sim(bxy=0.0, out_index = 1:3000, buy=0.4)$mr %>% mutate(overlap="disc_overlap", bxy=0.0, buy=0.4),
    gwas_sim(bxy=0.0, out_index = 3001:6000, buy=0.4)$mr %>% mutate(overlap="rep_overlap", bxy=0.0, buy=0.4),
    gwas_sim(bxy=0.0, out_index = 6001:9000, buy=0.4)$mr %>% mutate(overlap="no_overlap", bxy=0.0, buy=0.4),
    gwas_sim(bxy=0.0, out_index = 1501:4500, buy=0.4)$mr %>% mutate(overlap="both_overlap", bxy=0.0, buy=0.4),
    gwas_sim(bxy=0.2, out_index = 1:3000, buy=-0.4)$mr %>% mutate(overlap="disc_overlap", bxy=0.2, buy=-0.4),
    gwas_sim(bxy=0.2, out_index = 3001:6000, buy=-0.4)$mr %>% mutate(overlap="rep_overlap", bxy=0.2, buy=-0.4),
    gwas_sim(bxy=0.2, out_index = 6001:9000, buy=-0.4)$mr %>% mutate(overlap="no_overlap", bxy=0.2, buy=-0.4),
    gwas_sim(bxy=0.2, out_index = 1501:4500, buy=-0.4)$mr %>% mutate(overlap="both_overlap", bxy=0.2, buy=-0.4),
    gwas_sim(bxy=0.2, out_index = 1:3000, buy=0)$mr %>% mutate(overlap="disc_overlap", bxy=0.2, buy=0),
    gwas_sim(bxy=0.2, out_index = 3001:6000, buy=0)$mr %>% mutate(overlap="rep_overlap", bxy=0.2, buy=0),
    gwas_sim(bxy=0.2, out_index = 6001:9000, buy=0)$mr %>% mutate(overlap="no_overlap", bxy=0.2, buy=0),
    gwas_sim(bxy=0.2, out_index = 1501:4500, buy=0)$mr %>% mutate(overlap="both_overlap", bxy=0.2, buy=0),
    gwas_sim(bxy=0.2, out_index = 1:3000, buy=0.4)$mr %>% mutate(overlap="disc_overlap", bxy=0.2, buy=0.4),
    gwas_sim(bxy=0.2, out_index = 3001:6000, buy=0.4)$mr %>% mutate(overlap="rep_overlap", bxy=0.2, buy=0.4),
    gwas_sim(bxy=0.2, out_index = 6001:9000, buy=0.4)$mr %>% mutate(overlap="no_overlap", bxy=0.2, buy=0.4),
    gwas_sim(bxy=0.2, out_index = 1501:4500, buy=0.4)$mr %>% mutate(overlap="both_overlap", bxy=0.2, buy=0.4)
  )
}, mc.cores=10) %>% bind_rows()
```

Create figure

```{r}
ress <- filter(res) %>%
  tidyr::separate(what, into=c("exposure_dataset", "instruments"), "_") %>%
  group_by(exposure_dataset, instruments, overlap, buy, bxy) %>%
  summarise( n=n(), z=mean(abs(b/se)),prop_sig = sum(pnorm(abs(b/se), lower.tail=FALSE) < 0.05) / n(), se=sd(b)/sqrt(n),b=mean(b))




# Plot beta
p1 <- ggplot(subset(ress, bxy==0), aes(x=overlap, y=b, fill=exposure_dataset)) +
geom_bar(stat="identity", position="dodge", aes(fill=exposure_dataset)) +
geom_errorbar(aes(ymin = b - se, ymax = b + se), width=0.1,position=position_dodge(0.9)) +
facet_grid(instruments ~ buy) +
scale_fill_brewer(type="qual") +
geom_hline(yintercept=0, linetype="dotted") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
ggsave(p1, file=paste0(imgdir,"/replication_beta_0.png"), width=10, height=10)

p1 <- ggplot(subset(ress, bxy==0.2), aes(x=overlap, y=b, fill=exposure_dataset)) +
geom_bar(stat="identity", position="dodge", aes(fill=exposure_dataset)) +
geom_errorbar(aes(ymin = b - se, ymax = b + se), width=0.1,position=position_dodge(0.9)) +
facet_grid(instruments ~ buy) +
scale_fill_brewer(type="qual") +
geom_hline(yintercept=0.2, linetype="dotted") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
ggsave(p1, file=paste0(imgdir,"/replication_beta_0.2.png"), width=10, height=10)

# Plot proportion of significant p vals
p1 <- ggplot(subset(ress, bxy==0), aes(x=overlap, y=prop_sig, fill=exposure_dataset)) +
geom_bar(stat="identity", position="dodge", aes(fill=exposure_dataset)) +
facet_grid(instruments ~ buy) +
scale_fill_brewer(type="qual") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
ggsave(p1, file=paste0(imgdir,"/replication_prop_sig_0.png"), width=10, height=10)

p1 <- ggplot(subset(ress, bxy==0.2), aes(x=overlap, y=prop_sig, fill=exposure_dataset)) +
geom_bar(stat="identity", position="dodge", aes(fill=exposure_dataset)) +
facet_grid(instruments ~ buy) +
scale_fill_brewer(type="qual") +
theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))
ggsave(p1, file=paste0(imgdir,"/replication_prop_sig_0.2.png"), width=10, height=10)
```
